---
title: "xcms3 benchmarking"
output:
  BiocStyle::html_document2:
    toc: false
vignette: >
  % \VignetteIndexEntry{xcms3 benchmarking}
  % \VignetteEngine{knitr::rmarkdown}
  % \VignetteKeyword{metabolomics, mass spectrometry}
  % \VignettePackage{xcms}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

# Introduction

In this vignette we evaluate the performance of some recently added/modified
functionality of the `xcms` package and compare it, where possible, with the
*old* functions and methods from the `xcms` package.

```{r libs, message = FALSE}
library(xcms)
```

# Data binning

`xcms` uses binning of the MS data to generate the *profile* matrix on which
e.g. the `findPeaks.matchedFilter` method identified the MS1 features. Different
binning methods are available, such as `profBin`, that bins M/Z values into
equal-sized bins and identifies for all M/Z values falling into each bin the
maximal intensity value. This concept of data binning has been generalized and
is provided by the `binYonX` function that allows to bin any given numeric
vector. The function calls C-routines to ensure a high performance. In the
example below we compare the speed of the `profBin` and `binYonX` functions.

```{r binYonXbenchPrepare}
## Define a large numeric vector of sorted values.
X <- sort(abs(rnorm(500000, mean = 500, sd = 100)))

```

Bin this vector into 1000 bins using `profBin`.

```{r runProfBin}
system.time(a <- xcms:::profBin(X, X, 1000))
```

And using the `binYonX` function.

```{r runBinYonX}
system.time(b <- xcms:::binYonX_max(X, nBins = 1000, sortedX = TRUE,
                                    shiftByHalfBinSize = TRUE))
```

