---
title: "LC-MS feature grouping"
package: xcms
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{LC-MS feature grouping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{xcms,msdata,BiocStyle,faahKO,pheatmap,MsFeatures}
  %\VignettePackage{xcms}
  %\VignetteKeywords{mass spectrometry, metabolomics}
---

```{r biocstyle, echo = FALSE, results = "asis"}
BiocStyle::markdown()
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

**Package**: `r Biocpkg("xcms")`<br />
**Authors**: Johannes Rainer<br />
**Modified**: `r file.info("LC-MS-feature-grouping.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r init, results = "hide", echo = FALSE}
## Silently loading all packages
library(BiocStyle)
library(xcms)
library(MsFeatures)
register(SerialParam())

```

# Introduction

In a typical LC-MS-based metabolomics experiment compounds eluting from the
chromatography are first ionized before being measured by mass spectrometry
(MS). During the ionization different (multiple) ions can be generated from the
same compound which all will be measured by MS. In general, the resulting data
is then pre-processed to identify chromatographic peaks in the data and to group
these across samples in the correspondence analysis. The result are distinct
LC-MS features, characterized by their specific m/z and retention time
range. Different ions generated during ionization will be detected as different
features. To reduce data set complexity (and to aid in subsequent annotation
steps) it is advisable to group features which are supposedly representing
signal from the same original ion. This document describes functionality to aid
in this feature grouping step which are provided by this package.

General MS feature grouping concepts and functionality are provided by
Bioconductor's `r Biocpkg("MsFeatures")` package. Within `xcms` this
functionality is re-used and expanded to also allow feature grouping based on
similarity between the features' extracted ion chromatograms (EIC).


# LC-MS feature grouping

We demonstrate the feature grouping functionality on the simple toy data set
used also in the `r Biocpkg("xcms")` package and provided through the `faahKO`
package. This data set consists of samples from 4 mice with knock-out of the
fatty acid amide hydrolase (FAAH) and 4 wild type mice. Pre-processing of this
data set is described in detail in the *xcms* vignette of the `xcms`
package. Below we load all required packages and the result from this
pre-processing updating also the location of the respective raw data files on
the current machine.

```{r load-data}
library(xcms)
library(faahKO)
library(MsFeatures)

data("xdata")
## Update the path to the files for the local system
dirname(xdata) <- c(rep(system.file("cdf", "KO", package = "faahKO"), 4),
                    rep(system.file("cdf", "WT", package = "faahKO"), 4))
```

Before performing the feature grouping we inspect the result object. With
`featureDefinitions` we can extract the results from the correspondence
analysis.

```{r fdev}
featureDefinitions(xdata)
```

Each row in this data frame represents the definition of one feature, with its
average and range of m/z and retention time. Column `"peakidx"` provides the
index of each chromatographic peak which is assigned to the feature in the
`chromPeaks` matrix of the result object. The `featureValues` function allows to
extract *feature values*, i.e. a matrix with feature abundances, one row per
feature and columns representing the samples of the present data set.

Below we extract the feature values with and without *filled-in* peak
data. Without the gap-filled data only abundances from **detected**
chromatographic peaks are reported. In the gap-filled data, for samples in which
no chromatographic peak for a feature was detected, all signal from the m/z -
retention time range defined based on the detected chromatographic peaks was
integrated.
