#+TITLE: New and modified functionality in xcms

---
title: "New and modified functionality in xcms"
author: "Johannes Rainer"
graphics: yes
package: xcms
output:
  BiocStyle::html_document2:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{New and modified functionality in xcms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{xcms,RColorBrewer}
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---


# New functionality in `xcms`

This document describes new functionality and changes to existing functionality
in the `xcms` package introduced during the update to version *3*.

```{r message = FALSE, warning = FALSE}
library(xcms)
library(RColorBrewer)
```


## Modernized user interface

The modernization of the user interface comprises new classes for data
representation and new data analysis methods. In addition, the core logic for
the data processing has been extracted from the old methods and put into a set
of R functions, the so called core API functions (or `do_` functions), that take
standard R data structures as input and can hence also easily be included in
other R packages. The new user interface aims at simplifying and streamlining
the `xcms` workflow while guaranteeing data integrity and performance also for
large scale metabolomics experiments. All objects in the new user interface
ensuring data integrity *via* validation methods and class versioning, all methods
are tested internally in extensive unit tests to guarantee proper functionality.

In the new user interface, objects from the `MSnbase` Bioconductor package are
used, namely the `OnDiskMSnExp` object. This object is specifically designed for
large scale experiments by just loading basic information from the raw files and
reading the actual spectra data (mz and intensity values) on demand from the
files. In contrast to the old `xcmsRaw` objects this object contains information
from all files of an experiment. In addition, by using this object from the
`MSnbase` package as container for the raw data, data normalization and adjustment
methods defined for this class can be applied directly without having them to be
re-implemented in the `xcms` package. In the new user interface the results from
the pre-processing steps are stored in a new results object, the `XCMSnExp`
object. This is similar to the old `xcmsSet` object, but by extending the
`OnDiskMSnExp` object it inherits all of its methods and hence provides a simple
and streamlined access to the (raw) data on which the processing was performed.

For feature detection (peak calling), and thus replacing the `findPeaks.*`
methods, the `detectFeatures` method has been implemented.  The reason for
changing the name from `findPeaks` to `detectFeatures` was to discriminate *peak*
(i.e. mass peak mostly used in proteomics), from *feature* (which describes a peak
in chromatographic space). The `detectFeatures` method performs dispatching to
the various algorithms (e.g. *centWave*, *matchedFilter*) based on a parameter class
submitted *via* the `param` argument. This encapsulation of the parameters to a
function into a parameter class (such as `CentWaveParam`) avoids busy function
calls (with many single parameters to specify) and enables saving, reloading and
reusing settings as well as, by adding them to the result objects, enabling the
documentation of the data processing history. In addition, parameter validation
can be performed within the parameter object and hence is no longer required in
the analysis function.

The implementation of the `detectFeatures` method for `OnDiskMSnExp` and `MSnExp`
objects (from the `MSnbase` package) enable feature detection on these types of
objects that represent the full MS data of an experiment along with all
experiment relevant metadata. Loading the files first into an `MSnExp` or
`OnDiskMSnExp` (the latter being specifically designed for large scale
experiments) enables a first data inspection and quality assessment, followed by
eventual subsetting and filtering (e.g. on retention time) before performing the
feature detection on the thus processed data.

Below we load the raw data files from the `faahKO` package as an `OnDiskMSnExp`
object using the `readMSData2` function from the `MSnbase` package.

```{r message = FALSE, warning = FALSE}
## Reading the raw data using the MSnbase package
library(MSnbase)
library(xcms)
## Load the CDF files from the faahKO
cdf_files <- dir(system.file("cdf", package = "faahKO"), recursive = TRUE,
		 full.names = TRUE)

## Define the sample grouping.
s_groups <- rep("KO", length(cdf_files))
s_groups[grep(cdf_files, pattern = "WT")] <- "WT"
## Define a data.frame that will be used as phenodata
pheno <- data.frame(sample_name = sub(basename(cdf_files), pattern = ".CDF",
				      replacement = "", fixed = TRUE),
		    sample_group = s_groups, stringsAsFactors = FALSE)

## Read the data.
raw_data <- readMSData2(cdf_files, pdata = new("NAnnotatedDataFrame", pheno))
```

We next plot the total ion chromatogram (TIC) for all files within the
experiment. Note that we are iteratively sub-setting the full data per file,
which for `OnDiskMSnExp` is an efficient way to subset the data while ensuring
that all data, including metadata, stays consistent.

```{r faahKO-tic, message = FALSE, fig.align = 'center', fig.width = 8, fig.height = 4}
library(RColorBrewer)
sample_colors <- brewer.pal(3, "Set1")[1:2]
names(sample_colors) <- c("KO", "WT")
## Subset the full raw data by file and plot the data.
tmp <- filterFile(raw_data, file = 1)
plot(x = rtime(tmp), y = tic(tmp), xlab = "retention time", ylab = "TIC",
     col = sample_colors[pData(tmp)$sample_group], type = "l")
for (i in 2:length(fileNames(raw_data))) {
    tmp <- filterFile(raw_data, file = i)
    points(rtime(tmp), tic(tmp), type = "l",
	   col = sample_colors[pData(tmp)$sample_group])
}
legend("topleft", col = sample_colors, legend = names(sample_colors), lty = 1)
```

-   Do the feature detection.
-   Describe the feature detection methods.


## New naming convention

Methods for data analysis from the original `xcms` code have been renamed to avoid
potential confusions:

-   `detectFeatures` instead of `findPeaks`: for new functions and methods the term
    *peak* is avoided as much as possible, as it is usually used to describe a mass
    peak in mz dimension. To clearly distinguish between these peaks and peaks in
    retention time space, the latter are referred to as *features*.

-   `groupFeatures` instead of `group` to clearly indicate what is being
    grouped. Group might be a sample group or a feature group, hence groups of
    features are always referred to as *feature groups*.

-   `adjustRtime` instead of `retcor` for retention time correction. The word *cor* in
    *retcor* might be easily misinterpreted as *correlation* instead of correction.


## New data classes


### `OnDiskMSnExp`

This object is defined and documented in the `MSnbase` package. In brief, it is a
container for the full raw data from an MS-based experiment. To keep the memory
footprint low the mz and intensity values are only loaded from the raw data
files when required. The `OnDiskMSnExp` object replaces the `xcmsRaw` object.


### `XCMSnExp`

The `XCMSnExp` class extends the `OnDiskMSnExp` object from the `MSnbase` package and
represents a container for the xcms-based preprocessing results while (since it
inherits all functionality from its parent class) keeping a direct relation to
the (raw) data on which the processing was performed. An additional slot
`.processHistory` in the object allows to keep track of all performed processing
steps. Each analysis method, such as `detectFeatures` adds an `XProcessHistory`
object which includes also the parameter class passed to the analysis
method. Hence not only the time and type of the analysis, but its exact settings
are reported within the `XCMSnExp` object. The `XCMSnExp` is thus equivalent to the
`xcmsSet` from the original `xcms` implementation.


## Binning and missing value imputation functions

The binning/profile matrix generation functions have been completely
rewritten. The new `binYonX` function replaces the binning of intensity values
into bins defined by their m/z values implemented in the `profBin`, `profBinLin` and
`profBinLinBase` methods. The `binYonX` function provides also additional functionality:

-   Breaks for the bins can be defined based on either the number of desired bins
    (`nBins`) or the size of a bin (`binSize`). In addition it is possible to provide
    a vector with pre-defined breaks. This allows to bin data from multiple files
    or scans on the same bin-definition.

-   The function returns a list with element `y` containing the binned values and
    element `x` the bin mid-points.

-   Values in input vector `y` can be aggregated within each bin with different
    methods: `max`, `min`, `sum` and `mean`.

-   The index of the largest (or smallest for `method` being "min") within each bin
    can be returned by setting argument `returnIndex` to `TRUE`.

-   Binning can be performed on single or multiple sub-sets of the input vectors
    using the `fromIdx` and `toIdx` arguments. This replaces the *M* methods (such as
    `profBinM`). These sub-sets can be overlapping.

The missing value imputation logic inherently build into the `profBinLin` and
`profBinLinBase` methods has been implemented in the `imputeLinInterpol` function.

The example below illustrates the binning and imputation with the `binYtoX` and
`imputeLinInterpol` functions. After binning of the test vectors below some of the
bins have missing values, for which we impute a value using
`imputeLinInterpol`. By default, `binYonX` selects the largest value within each
bin, but other aggregation methods are also available (i.e. min, max, mean,
sum).

```{r message = FALSE}
## Defining the variables:
set.seed(123)
X <- sort(abs(rnorm(30, mean = 20, sd = 25))) ## 10
Y <- abs(rnorm(30, mean = 50, sd = 30))

## Bin the values in Y into 20 bins defined on X
res <- binYonX(X, Y, nBins = 22)

res
```

As a result we get a `list` with the bin mid-points (`$x`) and the binned `y` values
(`$y`).

Next we use two different imputation approaches, a simple linear interpolation
and the linear imputation approach that was defined in the `profBinLinBase`
method. The latter performs linear interpolation only considering a certain
neighborhood of missing values otherwise replacing the `NA` with a base value.

```{r binning-imputation-example, message = FALSE, fig.width = 10, fig.height = 7, fig.cap = 'Binning and missing value imputation results. Black points represent the input values, red the results from the binning and blue and green the results from the imputation (with method lin and linbase, respectively).'}
## Plot the actual data values.
plot(X, Y, pch = 16, ylim = c(0, max(Y)))
## Visualizing the bins
abline(v = breaks_on_nBins(min(X), max(X), nBins = 22), col = "grey")

## Define colors:
point_colors <- paste0(brewer.pal(4, "Set1"), 80)
## Plot the binned values.
points(x = res$x, y = res$y, col = point_colors[1], pch = 15)

## Perform the linear imputation.
res_lin <- imputeLinInterpol(res$y)

points(x = res$x, y = res_lin, col = point_colors[2], type = "b")

## Perform the linear imputation "linbase"
res_linbase <- imputeLinInterpol(res$y, method = "linbase")
points(x = res$x, y = res_linbase, col = point_colors[3], type = "b", lty = 2)
```

The difference between the linear interpolation method `lin` and `linbase` is that
the latter only performs the linear interpolation in a pre-defined neighborhood
of the bin with the missing value (`1` by default). The other missing values are
set to a base value corresponding to half of the smallest bin value. Both
methods thus yield same results, except for bins 15-17 (see Figure above).


## Core feature detection functions

The core logic from the feature detection methods `findPeaks.centWave`,
`findPeaks.massifquant`, `findPeaks.matchedFilter` has been extracted and put into
functions with the common prefix `do_detectFeatures` with the aim, as detailed in
issue [#30](https://github.com/sneumann/xcms/issues/30), to separate the core logic from the analysis methods invoked by the
users to enable also the use of the feature detection functions using base R
parameters (i.e. without specific classes containing the data such as the
`xcmsRaw` class). This simplifies also the re-use of these functions in other
packages and simplifies the future implementation of the feature detection
algorithms for e.g. the `MSnExp` or `OnDiskMSnExp` objects from the `MSnbase`
Bioconductor package. The implemented functions are:

-   `do_detectFeatures_centWave`: peak density and wavelet based feature detection
    for high resolution LC/MS data in centroid mode [@Tautenhahn:2008fx].
-   `do_detectFeatures_matchedFilter`: identification of features in the
    chromatographic domain based on matched filtration [@Smith:2006ic].
-   `do_detectFeatures_massifquant`: identification of features using Kalman
    filters.

One possible drawback from the introduction of this new layer is, that more
objects get copied by R which *could* eventually result in a larger memory demand
or performance decrease (while no such was decrease was observed up to now).


## Usability improvements

-   `[` subsetting method for `xcmsRaw` objects that enables to subset an `xcmsRaw`
    object to specific scans/spectra.
-   `profMat` method to extract the *profile* matrix from the `xcmsRaw` object. This
    method should be used instead of directly accessing the `@env$profile` slot, as
    it will create the profile matrix on the fly if it was not pre-calculated (or
    if profile matrix generation settings have been changed).


# Changes due to bug fixes and modified functionality


## Differences in linear interpolation of missing values (`profBinLin`).

From `xcms` version 1.51.1 on the new binning functions are used, thus, the bug
described here are fixed.

Two bugs are present in the `profBinLin` method (reported as issues [#46](https://github.com/sneumann/xcms/issues/46) and [#49](https://github.com/sneumann/xcms/issues/49) on
github) which are fixed in the new `binYonX` and `imputeLinInterpol` functions:

-   The first bin value calculated by `profBinLin` can be wrong (i.e. not being the
    max value within that bin, but the first).
-   If the last bin contains also missing values, the method fails to determine
    a correct value for that bin.

The `profBinLin` method is used in `findPeaks.matchedFilter` if the profile
method is set to "binlin".

The example below illustrates both differences.

```{r }
## Define a vector with empty values at the end.
X <- 1:11
set.seed(123)
Y <- sort(rnorm(11, mean = 20, sd = 10))
Y[9:11] <- NA
nas <- is.na(Y)
## Do interpolation with profBinLin:
resX <- xcms:::profBinLin(X[!nas], Y[!nas], 5, xstart = min(X),
			  xend = max(X))
resX
res <- binYonX(X, Y, nBins = 5L, shiftByHalfBinSize = TRUE)
resM <- imputeLinInterpol(res$y, method = "lin",
			  noInterpolAtEnds = TRUE)
resM
```

Plotting the results helps to better compare the differences. The black points
in the figure below represent the actual values of `Y` and the grey vertical lines
the breaks defining the bins. The blue lines and points represent the result
from the `profBinLin` method. The bin values for the first and 4th bin are clearly
wrong. The green colored points and lines represent the results from the `binYonX`
and `imputeLinInterpol` functions (showing the correct binning and interpolation).

```{r profBinLin-problems, message = FALSE, fig.align = 'center', fig.width=10, fig.height = 7, fig.cap = "Illustration of the two bugs in profBinLin. The input values are represented by black points, grey vertical lines indicate the bins. The results from binning and interpolation with profBinLin are shown in blue and those from binYonX in combination with imputeLinInterpol in green."}
plot(x = X, y = Y, pch = 16, ylim = c(0, max(Y, na.rm = TRUE)),
     xlim = c(0, 12))
## Plot the breaks
abline(v = breaks_on_nBins(min(X), max(X), 5L, TRUE), col = "grey")
## Result from profBinLin:
points(x = res$x, y = resX, col = "blue", type = "b")
## Results from imputeLinInterpol
points(x = res$x, y = resM, col = "green", type = "b",
       pch = 4, lty = 2)
```

Note that by default `imputeLinInterpol` would also interpolate missing values at
the beginning and the end of the provided numeric vector. This can be disabled
(to be compliant with `profBinLin`) by setting parameter `noInterpolAtEnds` to
`TRUE` (like in the example above).


## Differences due to updates in `do_detectFeatures_matchedFilter`, respectively `findPeaks.matchedFilter`.

The original `findPeaks.matchedFilter` (up to version 1.49.7) had several
shortcomings and bugs that have been fixed in the new
`do_detectFeatures_matchedFilter` method:

-   The internal iterative processing of smaller chunks of the full data (also
    referred to as *iterative buffering*) could result, for some bin (step) sizes to
    unstable binning results (discussed in issue [#47](https://github.com/sneumann/xcms/issues/47) on github): calculation of
    the breaks, or to be precise, the actually used bin size was performed in each
    iteration and could lead to slightly different sizes between iterations (due
    to rounding errors caused by floating point number representations in C).

-   The iterative buffering raises also a conceptual issue when linear
    interpolation is performed to impute missing values: the linear imputation
    will only consider values within the actually processed buffer and can thus
    lead to wrong or inaccurate imputations.

-   The `profBinLin` implementation contains two bugs, one that can result in
    failing to identify the maximal value in the first and last bin (see issue
    [#46](https://github.com/sneumann/xcms/issues/46)) and one that fails to assign a value to a bin (issue [#49](https://github.com/sneumann/xcms/issues/49)). Both are fixed
    in the `do_detectFeatures_matchedFilter` implementation.

A detailed description of tests comparing all implementations is available in
issue [#52](https://github.com/sneumann/xcms/issues/52) on github. Note also that in course of these changes also the `getEIC`
method has been updated to use the new binning and missing value imputation
function.

While it is strongly discouraged, it is still possible to use to *old* code (from
1.49.7) by calling `useOriginalCode(TRUE)`.


## Differences in `findPeaks.massifquant`

-   Argument `scanrange` was ignored in the *original* old code (issue [#61](https://github.com/sneumann/xcms/issues/61)).
-   The method returned a `matrix` if `withWave` was `0` and a `xcmsPeaks` object
    otherwise. The updated version returns **always** an `xcmsPeaks` object (issue #60).


## Differences in *obiwarp* retention time correction

Retention time correction using the obiwarp method uses the *profile* matrix
(i.e. intensities binned in discrete bins along the mz axis). Profile matrix
generation uses now the `binYonX` method which fixed some problems in the original
binning and linear interpolation methods. Thus results might be slightly
different.


## `retcor.peaksgroups`: change in the way how *well behaved* feature groups are ordered

The `retcor.peakgroups` defines first the feature groups (peaks groups) that are
used for the alignment of all spectra. Once these are identified, the retention
time of the feature with the highest intensity in a sample for a given feature
group is returned and the feature groups are ordered increasingly by retention
time (which is required for the later fitting of either a polynomial or a linear
model to the data). The selection of the retention time of the feature with the
highest intensity within a feature group and samples, denoted as *representative*
feature for a given feature group in a sample, ensures that only the retention
time of a single feature per sample and feature group is selected (multiple
features within the same sample can be assigned to a feature group).  In the
original code the ordering of the feature groups was however performed using the
median retention time of the complete feature group (which includes also
potential additional features per sample). This has been changed and the feature
groups are ordered now by the median retention time across samples of the
representative features.


## `scanrange` parameter in all `findPeaks` methods

The `scanrange` in the `findPeaks` methods is supposed to enable the peak detection
only within a user-defined range of scans. This was however not performed in
each method. Due to a bug in `findPeaks.matchedFilter`'s original code the
argument was ignored, except if the upper scan number of the user defined range
was larger than the total number of available scans (see issue [#63](https://github.com/sneumann/xcms/issues/63)). In
`findPeaks.massifquant` the argument was completely ignored (see issue [#61](https://github.com/sneumann/xcms/issues/61)) and,
while the argument was considered in `findPeaks.centWave` and feature detection
was performed within the specified scan range, but the original `@scantime` slot
was used throughout the code instead of just the scan times for the specified
scan indices (see issue [#64](https://github.com/sneumann/xcms/issues/64)).

These problems have been fixed in version 1.51.1 by first sub-setting the
`xcmsRaw` object (using the `[` method) before actually performing the feature
detection.


# Under the hood changes

These changes and updates will not have any large impact on the day-to-day use of
`xcms` and are listed here for completeness.

-   From `xcms` version 1.51.1 on the default methods from the `mzR` package are used
    for data import. Besides ensuring easier maintenance, this enables also data
    import from *gzipped* mzML files.


# Deprecated functions and files

Here we list all of the functions and related files that are deprecated.

-   `xcmsParallelSetup`, `xcmsPapply`, `xcmsClusterApply`: use `BiocParallel` package
    instead to setup and perform parallel processing, either *via* the `BPPARAM`
    parameter to function and methods, or by calling `register` to globally set
    parallel processing.

-   `profBin`, `profBinM`, `profBinLin`, `profBinLinM`, `profBinLinBase`, `profBinLinBaseM`:
    replaced by the `binYonX` and `imputeLinInterpol` functions. Also, to create or
    extract the profile matrix from an `xcmsRaw` object, the `profMat` method.


## Deprecated


### xcms 1.49:

-   `xcmsParallelSetup` (Deprecated.R)
-   `xcmsPapply` (Deprecated.R)
-   `xcmsClusterApply` (Deprecated.R)


### xcms 1.51:

-   `profBin` (c.R)
-   `profBinM` (c.R)
-   `profBinLin` (c.R)
-   `profBinLinM` (c.R)
-   `profBinLinBase` (c.R)
-   `profBinLinBaseM` (c.R)


## Defunct


# References

