---
title: "New and modified functionality in xcms"
graphics: yes
output:
  BiocStyle::html_document2:
    toc: false
vignette: >
  %\VignetteIndexEntry{New and modified functionality in xcms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{xcms,RColorBrewer}
  %\VignettePackage{xcms}
  %\VignetteKeywords{mass spectrometry, metabolomics}
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("xcms")`<br />
**Authors**: Johannes Rainer<br />
**Modified**: 20 October, 2016<br />
**Compiled**: `r date()`

# New functionality in `xcms`

This document describes new functionality and changes to existing functionality
in the `xcms` package introduced during the update to version *3*.

```{r message = FALSE}
library(xcms)
library(RColorBrewer)
```

## Binning and missing value imputation functions

The binning/profile matrix generation functions have been completely
rewritten. The new `binYonX` function replaces the binning of intensity values
into bins defined by their m/z values implemented in the `profBin`, `profBinLin` and
`profBinLinBase` methods. The `binYonX` function provides also additional functionality:

-   Breaks for the bins can be defined based on either the number of desired bins
    (`nBins`) or the size of a bin (`binSize`). In addition it is possible to provide
    a vector with pre-defined breaks. This allows to bin data from multiple files
    or scans on the same bin-definition.

-   The function returns a list with element `y` containing the binned values and
    element `x` the bin mid-points.

-   Values in input vector `y` can be aggregated within each bin with different
    methods: `max`, `min`, `sum` and `mean`.

-   The index of the largest (or smallest for `method` being "min") within each bin
    can be returned by setting argument `returnIndex` to `TRUE`.

-   Binning can be performed on single or multiple sub-sets of the input vectors
    using the `fromIdx` and `toIdx` arguments. This replaces the *M* methods (such as
    `profBinM`). These sub-sets can be overlapping.

The missing value imputation logic inherently build into the `profBinLin` and
`profBinLinBase` methods has been implemented in the `imputeLinInterpol` function.

The example below illustrates the binning and imputation with the `binYtoX` and
`imputeLinInterpol` functions. After binning of the test vectors below some of the
bins have missing values, for which we impute a value using
`imputeLinInterpol`. By default, `binYonX` selects the largest value within each
bin, but other aggregation methods are also available (i.e. min, max, mean,
sum).

```{r message = FALSE}
## Defining the variables:
set.seed(123)
X <- sort(abs(rnorm(30, mean = 20, sd = 25))) ## 10
Y <- abs(rnorm(30, mean = 50, sd = 30))

## Bin the values in Y into 20 bins defined on X
res <- binYonX(X, Y, nBins = 22)

res
```

As a result we get a `list` with the bin mid-points (`$x`) and the binned `y` values
(`$y`).

Next we use two different imputation approaches, a simple linear interpolation
and the linear imputation approach that was defined in the `profBinLinBase`
method. The latter performs linear interpolation only considering a certain
neighborhood of missing values otherwise replacing the `NA` with a base value.

```{r binning-imputation-example, message = FALSE, fig.width = 10, fig.height = 7, fig.cap = 'Binning and missing value imputation results. Black points represent the input values, red the results from the binning and blue and green the results from the imputation (with method lin and linbase, respectively).'}
## Plot the actual data values.
plot(X, Y, pch = 16, ylim = c(0, max(Y)))
## Visualizing the bins
abline(v = breaks_on_nBins(min(X), max(X), nBins = 22), col = "grey")

## Define colors:
point_colors <- paste0(brewer.pal(4, "Set1"), 80)
## Plot the binned values.
points(x = res$x, y = res$y, col = point_colors[1], pch = 15)

## Perform the linear imputation.
res_lin <- imputeLinInterpol(res$y)

points(x = res$x, y = res_lin, col = point_colors[2], type = "b")

## Perform the linear imputation "linbase"
res_linbase <- imputeLinInterpol(res$y, method = "linbase")
points(x = res$x, y = res_linbase, col = point_colors[3], type = "b", lty = 2)
```

The difference between the linear interpolation method `lin` and `linbase` is that
the latter only performs the linear interpolation in a pre-defined neighborhood
of the bin with the missing value (`1` by default). The other missing values are
set to a base value corresponding to half of the smallest bin value. Both
methods thus yield same results, except for bins 15-17 (see Figure above).

## Core feature detection functions

The core logic from the feature detection methods `findPeaks.centWave`,
`findPeaks.massifquant`, `findPeaks.matchedFilter` has been extracted and put into
functions with the common prefix `do_detectFeatures` with the aim, as detailed in
issue [#30](https://github.com/sneumann/xcms/issues/30), to separate the core logic from the analysis methods invoked by the
users to enable also the use of the feature detection functions using base R
parameters (i.e. without specific classes containing the data such as the
`xcmsRaw` class). This simplifies also the re-use of these functions in other
packages. The implemented functions are:

-   `do_detectFeatures_centWave`: peak density and wavelet based feature detection
    for high resolution LC/MS data in centroid mode [@Tautenhahn:2008fx].
-   `do_detectFeatures_matchedFilter`: identification of features in the
    chromatographic domain based on matched filtration [@Smith:2006ic].
-   `do_detectFeatures_massifquant`: iden

On possible drawback from the introduction of this new layer is, that more
objects get copied by R which *could* eventually result in a larger memory demand
or performance decrease (while no such was decrease was observed up to now).

## Usability improvements

-   `[` subsetting method for `xcmsRaw` objects that enables to subset an `xcmsRaw`
      object to specific scans/spectra.

# Changes due to bug fixes and modified functionality

## Differences in linear interpolation of missing values (`profBinLin`).

Two bugs are present in the `profBinLin` method (reported as issues [#46](https://github.com/sneumann/xcms/issues/46) and [#49](https://github.com/sneumann/xcms/issues/49) on
github) which are fixed in the new `binYonX` and `imputeLinInterpol` functions:

-   The first bin value calculated by `profBinLin` can be wrong (i.e. not being the
    max value within that bin, but the first).
-   If the last bin contains also missing values, the method fails to determine
    a correct value for that bin.

The `profBinLin` method is used in `findPeaks.matchedFilter` if the profile
method is set to "binlin".

The example below illustrates both differences.

```{r }
## Define a vector with empty values at the end.
X <- 1:11
set.seed(123)
Y <- sort(rnorm(11, mean = 20, sd = 10))
Y[9:11] <- NA
nas <- is.na(Y)
## Do interpolation with profBinLin:
resX <- xcms:::profBinLin(X[!nas], Y[!nas], 5, xstart = min(X),
			  xend = max(X))
resX
res <- binYonX(X, Y, nBins = 5L, shiftByHalfBinSize = TRUE)
resM <- imputeLinInterpol(res$y, method = "lin",
			  noInterpolAtEnds = TRUE)
resM
```

Plotting the results helps to better compare the differences. The black points
in the figure below represent the actual values of `Y` and the grey vertical lines
the breaks defining the bins. The blue lines and points represent the result
from the `profBinLin` method. The bin values for the first and 4th bin are clearly
wrong. The green colored points and lines represent the results from the `binYonX`
and `imputeLinInterpol` functions (showing the correct binning and interpolation).

```{r profBinLin-problems, message = FALSE, fig.align = 'center', fig.width=10, fig.height = 7, fig.cap = "Illustration of the two bugs in profBinLin. The input values are represented by black points, grey vertical lines indicate the bins. The results from binning and interpolation with profBinLin are shown in blue and those from binYonX in combination with imputeLinInterpol in green."}
plot(x = X, y = Y, pch = 16, ylim = c(0, max(Y, na.rm = TRUE)),
     xlim = c(0, 12))
## Plot the breaks
abline(v = breaks_on_nBins(min(X), max(X), 5L, TRUE), col = "grey")
## Result from profBinLin:
points(x = res$x, y = resX, col = "blue", type = "b")
## Results from imputeLinInterpol
points(x = res$x, y = resM, col = "green", type = "b",
       pch = 4, lty = 2)
```

Note that by default `imputeLinInterpol` would also interpolate missing values at
the beginning and the end of the provided numeric vector. This can be disabled
(to be compliant with `profBinLin`) by setting parameter `noInterpolAtEnds` to
`TRUE` (like in the example above).

## Differences due to updates in `do_detectFeatures_matchedFilter`, respectively `findPeaks.matchedFilter`.

The original `findPeaks.matchedFilter` (up to version 1.49.7) had several
shortcomings and bugs that have been fixed in the new
`do_detectFeatures_matchedFilter` method:

-   The internal iterative processing of smaller chunks of the full data (also
    referred to as *iterative buffering*) could result, for some bin (step) sizes to
    unstable binning results (discussed in issue [#47](https://github.com/sneumann/xcms/issues/47) on github): calculation of
    the breaks, or to be precise, the actually used bin size was performed in each
    iteration and could lead to slightly different sizes between iterations (due
    to rounding errors caused by floating point number representations in C).

-   The iterative buffering raises also a conceptual issue when linear
    interpolation is performed to impute missing values: the linear imputation
    will only consider values within the actually processed buffer and can thus
    lead to wrong or inaccurate imputations.

-   The `profBinLin` implementation contains two bugs, one that can result in
    failing to identify the maximal value in the first and last bin (see issue
    [#46](https://github.com/sneumann/xcms/issues/46)) and one that fails to assign a value to a bin (issue [#49](https://github.com/sneumann/xcms/issues/49)). Both are fixed
    in the `do_detectFeatures_matchedFilter` implementation.

A detailed description of tests comparing all implementations is available in
issue [#52](https://github.com/sneumann/xcms/issues/52) on github.

While it is strongly discouraged, it is still possible to use to *old* code (from
1.49.7) by calling `useOriginalCode(TRUE)`.

## Differences in `findPeaks.massifquant`

-   Argument `scanrange` was ignored in the *original* old code (issue [#61](https://github.com/sneumann/xcms/issues/61)).
-   The method returned a `matrix` if `withWave` was `0` and a `xcmsPeaks` object
    otherwise. The updated version returns **always** an `xcmsPeaks` object (issue #60).

## `scanrange` parameter in all `findPeaks` methods

The `scanrange` in the `findPeaks` methods is supposed to enable the peak detection
only within a user-defined range of scans. This was however not performed in
each method. Due to a bug in `findPeaks.matchedFilter`'s original code the
argument was ignored, except if the upper scan number of the user defined range
was larger than the total number of available scans (see issue [#63](https://github.com/sneumann/xcms/issues/63)). In
`findPeaks.massifquant` the argument was completely ignored (see issue [#61](https://github.com/sneumann/xcms/issues/61)) and,
while the argument was considered in `findPeaks.centWave` and feature detection
was performed within the specified scan range, but the original `@scantime` slot
was used throughout the code instead of just the scan times for the specified
scan indices (see issue [#64](https://github.com/sneumann/xcms/issues/64)).

These problems have been fixed in version 1.51.1 by first sub-setting the
`xcmsRaw` object (using the `[` method) before actually performing the feature
detection.

# Under the hood changes

These changes and update will not have any large impact on the day-to-day use of
`xcms` and are listed here for completeness.

# References
